Graphical Tests for Power Comparison of Competing Designs

### 2012-03-30 intro.tex
typos

Here we suggest a new approach using recent research on visual
inference \citep{buja:2009,wickham:2010}. In visual inference, real
data plots are considered to be test statistics, and these are
compared with plots of data generated from a null hypothesis using a
lineup. A lineup is similar to the police lineup, and this is where
the name comes from. The suspect (the plot of the real data) is placed
randomly in a field of plots of null data, and an impartial viewer is
asked to identify the plot that is the most different. If the viewer
identifies the suspect it lends statistical signficance to the
conclusion that the real data is not consistent with the null
data. Figure \ref{lineup} has an example of a lineup. Overlaid density
plots are used to display the distribution of two groups (blue, red)
of data. We are interested to know whether the real data has a
significant diffeerence between the centers of the two groups. Plot
numbered 13 contains the real data. Did you pick that one? In this
plot the blue group has a flat density whereas the density of red
group is more peaked and concentrated to the left of the blue.

@@

The previous work \citet{buja:2009,wickham:2010} describes the use of
lineups for evaluating discoveries made in exploratory analysis. Here
we describe the use of lineups for evalating plot design for
communicating findings visually. Different lineups are created, using
the same data, same null plots and positions within the lineup, but
different plot designs. Comparing how long and how accurately the
viewers take to report the feature of interest will assess which
design is better for the task. To investigate the benefit of lineups
for plot design evaluation we have conducted an experiment that also
contains simulated data, where the data is constructed to match and
expand on the features of interest in the real data. For example, in a
simple situation, where the discovered feature of interest is a
difference between the centers of two distributions, simulated data
would be constructed roughly matching the distribution of the real
data that varies the distance between centers, and also spread and
sample size. Lineups of these simulated data (embedded with simulated
null data plots) are generated and evaluated along with the lineups of
real data. The simulation study provides a backdrop for the real data,
which enables the broad applicability to be studied and a gauge for
the signal strength of the pattern in the real data.

@@

Lineups provide a way to statistically quantify the significance of the finding
\citet{majumder:2011}. Think of the linup as a set of $m$ plots, one of which is the real data. For one lineup and one observer, the significance level, $\alpha$ is $1/m$. This is the probability that the observer picks the plot of the real data by chance. It is the probability of picking the real data plot, when it is really *not* different from the other null plots, that is, a Type I error. (In Figure \ref{lineup}, $m=20$.) The power of a test is the probability that the real data plot is identified when it really is different from the null plots. Power is usually difficult to calculate because it is requires specific use of alternative hypothesis to calculate the probability.  Work in \citet{niladri:2012} addresses this to some extent with measures on the quality of a lineup, how different numerically, as best it can be calculated, the real data plot is from the null plots. 
%

When there are multiple observers the significance level also depends on the number of independent observers ($n$). Let $n$ be the number of independent observers and $x_i$ the number of observers who picked plot $i$, $i \in \{1, ..., m\}$, from the lineup. Then $(x_1, x_2, ..., x_m)$ follows a multinomial distribution Mult$_{\pi_1, \pi_2, ..., \pi_m}(x_1, x_2, ..., x_m)$ with $\sum_i \pi_i = 1$, where $\pi_i$ is the probability that plot $i$ is picked by an observer, which we can estimate as $\widehat{\pi_i} = x_i/n$.

*** Under the null that the real is no different from the null, isn't $\pi_i=1/m$? Fixing this then I think the next paragraph on signal strength flows right from this one. It is when we are calculating signal strength that we start counting the number of observers who pick it.

%The power of a test is the probability to reject the null hypothesis, when it is not true. In the framework of lineups, power is, effectively, the probability that the data plot is identified. , i.e. power can be estimated as $x/n$, where $x$ is the number of observers who correctly identifies the plot of the real data out of $n$ observers.

%Let $x$ be the number of observers who
%correctly identify the plot of the real data. The probability that one
%observer correctly identifies the real data plot, in a size $m$
%lineup, by chance is $1/m$.

\paragraph{Signal Strength}
To incorporate the results from multiple
observers think about the lineup as $m-1$ head-to-head competitions of
the data plot against $m-1$ null plots, which the data plot has to
win, is picked as the plot with the most distinct features. Assume
that the data plot is the one with the strongest signal, then the
probability that the data plot is picked by $x$ observers follows a
Beta distribution for the maximum \citet{****}:
\[
\text{signal strength} =  \left( \frac{x}{n}\right)^{1/(m-1)}.
\]
We call it signal strength \citet{majumder:2012}, and it is analogous to 1-$p$-value of a classical hypothesis test. Signal strength will range between 0 and 1, with higher values indicating stronger signal, more distinctive feature in the data.

The significance depends on the size ($m$) of
the lineup ($m=20$ in Figure \ref{lineup}) and the number of
independent observers ($n$). Let $x$ be the number of observers who
correctly identify the plot of the real data. The probability that one
observer correctly identifies the real data plot, in a size $m$
lineup, by chance is $1/m$. To incorporate the results from multiple
observers think about the lineup as $m-1$ head-to-head competitions of
the data plot against $m-1$ null plots, which the data plot has to
win, is picked as the plot with the most distinct features. Assume
that the data plot is the one with the strongest signal, then the
probability that the data plot is picked by $x$ observers follows a
Beta distribution for the maximum \citet{****}:
\[
\text{signal strength} =  \left( \frac{x}{n}\right)^{1/(m-1)}.
\]
We call it signal strength, and it is analgous to power in classical hypothesis testing. Signal strength will range between 0 and 1, with higher values indicating stronger signal, more distinctive feature in the data.

*** The above paragraph needs more explanation, and a reference for the beta distribution. I don't think the Majumder paper really is the source for this result is it? It leads up to it, but this is the new work that is not submitted yet.

*** Also need to refer to Niladri's ``Where's Waldo paper'' here too.

%The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup. Assume a lineup of size $m$ is evaluated by $n$ independent observers, out of which $x$ identify the data plot correctly, which we can use as an estimate for the probability to correctly identify the data plot: this leads us to an estimate for the probability of correctly identifying the plot of $x/n$, with 95\% confidence intervals  for $x/n$ given as $x/n \pm 1.96 \cdot \sqrt{x/n \cdot (1-x/n)}$ limited to the interval [0,1]. Note, that the resulting interval is often too large to be useful.

%We can think of a lineup as $m-1$ head-to-head comparisons of the dataplot and $m-1$ nullplots, which the dataplot all has to `win', in order to be picked as the plot with the most distinct features. Assuming that observers identify the plot with the strongest signal, the probability that data plot is picked follows a Beta distribution for the maximum: this allows us to introduce the {\it signal strength} \cite{majumder:2011} of a lineup as:

%Signal strength is limited to a range of values between 0 and 1. Higher values indicate stronger signal. In the case of a parametric situation (i.e. the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), it can be shown \cite{majumder:2011} that the signal strength is strongly correlated to the $p$-value of the corresponding statistical test for $\theta$.

Signal strength is used in our new work to evaluate the effectiveness of different plot designs. This paper describes two examples where we had real data analysis problems and decisions to make in order to communicate the results. Section \ref{designs} describes the process for comparing designs. Section \ref{setup} and describe the two data analysis problems and the experiments conducted to evaluate the plots designs. Section \ref{results} presents the findings of the experiments, and Section \ref{conclusions} presents conclusions and suggestions for future use.

>>>>

Here we suggest a new approach using recent research on visual
inference \citep{buja:2009,wickham:2010}. In visual inference, real
data plots are considered to be test statistics, and these are
compared with plots of data generated from a null hypothesis using a
lineup. A lineup is similar to the police lineup, and this is where
the name comes from. The suspect (the plot of the real data) is placed
randomly in a field of plots of null data, and an impartial viewer is
asked to identify the plot that is the most different. If the viewer
identifies the suspect it lends statistical significance to the
conclusion that the real data is not consistent with the null
data. Figure \ref{lineup} has an example of a lineup. Overlaid density
plots are used to display the distribution of two groups (blue, red)
of data. We are interested to know whether the real data has a
significant difference between the centers of the two groups. Plot
numbered 13 contains the real data. Did you pick that one? In this
plot the blue group has a flat density whereas the density of red
group is more peaked and concentrated to the left of the blue.

@@

The previous work \citet{buja:2009,wickham:2010} describes the use of
lineups for evaluating discoveries made in exploratory analysis. Here
we describe the use of lineups for evaluating plot design for
communicating findings visually. Different lineups are created, using
the same data, same null plots and positions within the lineup, but
different plot designs. Comparing how long and how accurately the
viewers take to report the feature of interest will assess which
design is better for the task. To investigate the benefit of lineups
for plot design evaluation we have conducted an experiment that also
contains simulated data, where the data is constructed to match and
expand on the features of interest in the real data. For example, in a
simple situation, where the discovered feature of interest is a
difference between the centers of two distributions, simulated data
would be constructed roughly matching the distribution of the real
data that varies the distance between centers, and also spread and
sample size. Lineups of these simulated data (embedded with simulated
null data plots) are generated and evaluated along with the lineups of
real data. The simulation study provides a backdrop for the real data,
which enables the broad applicability to be studied and a gauge for
the signal strength of the pattern in the real data.

@@

Lineups provide a way to statistically quantify the significance of the finding \citet{majumder:2011}: 
think of the lineup as a set of $m$ plots, one of which is the real data. 
The probability that an observer picks the real data plot just by chance, i.e. when it is really *not* different from the other null plots, is $1/m$. 
If an observer is able to identify the data plot from the lineup, we reject the null hypothesis that the null plot is not different from the data.  This sets  our Type I error rate  $\alpha$ at a level of $1/m$. (In Figure \ref{lineup}, $m=20$)

When there are multiple independent observers ($n$) we have more freedom in setting  the significance level: assume that $x$ out of those $n$ observers picked the data plot. Let $X$ be the corresponding random variable, i.e. $X$ = \# times out of $n$ independent repetitions that  the data is picked from the lineup. Under the null hypothesis, $X$ has a Binomial distribution: $X \sim B_{n, 1/m}$.
We can then compute the $p$ value of a lineup as the probability to have $x$ or more observers picking the data plot (under the assumption that the null hypothesis is true, i.e. the plot is not different):
\[
p\text{-value} = P(X \ge x \mid H_0) =   1 - B_{n, 1/m} (x-1).
\]

When comparing different tests of the same quantity, we consider that test better, that has greater {\it power}.
The power of a test is the probability to reject the null hypothesis -- in a lineup this is the probability that an observer identifies the plot of the real data. 
Analytically, power is usually difficult to calculate because it is requires specific use of an alternative hypothesis to calculate the probability.  Work in \citet{niladri:2012} addresses this to some extent with measures on the quality of a lineup, how different numerically, as best it can be calculated, the real data plot is from the null plots.

In our situation, though, it is fairly straightforward to {\it estimate} the power of a lineup:

Let $n$ be the number of independent observers and $x_i$ the number of observers who picked plot $i$, $i \in \{1, ..., m\}$, from the lineup. Then $(x_1, x_2, ..., x_m)$ follows a multinomial distribution Mult$_{\pi_1, \pi_2, ..., \pi_m}(x_1, x_2, ..., x_m)$ with $\sum_i \pi_i = 1$, where $\pi_i$ is the probability that plot $i$ is picked by an observer, which we can estimate as $\widehat{\pi_i} = x_i/n$.

The power of a lineup can therefore be estimated as the ratio of correct identifications $x$ out of $n$ viewings.

%*** Under the null that the real is no different from the null, isn't $\pi_i=1/m$? Fixing this then I think the next paragraph on signal strength flows right from this one. It is when we are calculating signal strength that we start counting the number of observers who pick it.
%
%When there are multiple observers the significance level also depends on the number of independent observers ($n$). 
%Note that under the null hypothesis - i.e. when the real data is not different from the others - we expect $\pi_i = 1/m$.
%
%
%%The power of a test is the probability to reject the null hypothesis, when it is not true. In the framework of lineups, power is, effectively, the probability that the data plot is identified. , i.e. power can be estimated as $x/n$, where $x$ is the number of observers who correctly identifies the plot of the real data out of $n$ observers.
%
%%Let $x$ be the number of observers who
%%correctly identify the plot of the real data. The probability that one
%%observer correctly identifies the real data plot, in a size $m$
%%lineup, by chance is $1/m$. 
%
%\paragraph{Signal Strength}
%To incorporate the results from multiple
%observers think about the lineup as $m-1$ head-to-head competitions of
%the data plot against $m-1$ null plots, which the data plot has to
%win, to be picked as the plot with the most distinct features. 
%Assuming that observers are picking the plot with the strongest signal,  the
%probability that the data plot is picked by $x$ observers follows a
%Beta distribution for the maximum \citet{****}, which we call {\it signal strength} $\Xi$ \citet{majumder:2012}:
%\[
%\Xi =  \left( \frac{x}{n}\right)^{1/(m-1)}.
%\]
%Signal strength will range between 0 and 1, with higher values indicating stronger signal, more distinctive feature in the data. $1-\Xi$ is analogous to the $p$-value of a classical hypothesis test. 
%
%
%
%
%*** The above paragraph needs more explanation, and a reference for the beta distribution. I don't think the Majumder paper really is the source for this result is it? It leads up to it, but this is the new work that is not submitted yet.
%
%*** Also need to refer to Niladri's ``Where's Waldo paper'' here too.
%
%%The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup. Assume a lineup of size $m$ is evaluated by $n$ independent observers, out of which $x$ identify the data plot correctly, which we can use as an estimate for the probability to correctly identify the data plot: this leads us to an estimate for the probability of correctly identifying the plot of $x/n$, with 95\% confidence intervals  for $x/n$ given as $x/n \pm 1.96 \cdot \sqrt{x/n \cdot (1-x/n)}$ limited to the interval [0,1]. Note, that the resulting interval is often too large to be useful.
%
%%We can think of a lineup as $m-1$ head-to-head comparisons of the dataplot and $m-1$ nullplots, which the dataplot all has to `win', in order to be picked as the plot with the most distinct features. Assuming that observers identify the plot with the strongest signal, the probability that data plot is picked follows a Beta distribution for the maximum: this allows us to introduce the {\it signal strength} \cite{majumder:2011} of a lineup as:
%
%%Signal strength is limited to a range of values between 0 and 1. Higher values indicate stronger signal. In the case of a parametric situation (i.e. the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), it can be shown \cite{majumder:2011} that the signal strength is strongly correlated to the $p$-value of the corresponding statistical test for $\theta$.
%

The power of lineups is used in our new work to evaluate the effectiveness of different plot designs. This paper describes two examples where we had real data analysis problems and decisions to make in order to communicate the results. Section \ref{designs} describes the process for comparing designs. Section \ref{setup} and describe the two data analysis problems and the experiments conducted to evaluate the plots designs. Section \ref{results} presents the findings of the experiments, and Section \ref{conclusions} presents conclusions and suggestions for future use.
